---
title: "Most excellent figures in R with ggplot2"
author: "Ben Williams"
date: "2020-05"
output: 
   html_document:
    toc: true
    toc_depth: 2
    theme: cerulean
    fig_caption: true
---
```{r, echo=F}
knitr::opts_chunk$set(warning = F, message=F, error=F)
```

# 1 Short version  
Set up the R work environment to produce publication quality documents using ggplot.
Use `theme_set` to define the base figure elements - in this case: no gridlines, a white background, and 12 pt times new roman font.

*Note: this assumes you are working in a Windows environment* 

Load packages

```{r theme_short, eval=F}
library(extrafont)
#font_import() only do this one time - it takes a while
loadfonts(device="win")
windowsFonts(Times=windowsFont("TT Times New Roman"))

library(ggplot2)
theme_set(theme_bw(base_size=12, base_family='Times New Roman')+ 
  theme(panel.grid.major = element_blank(),
  panel.grid.minor = element_blank()))
```

Use `ggsave` to increase dpi from base R level (72 dpi) and to defining figure dimensions.

```{r, eval=F}
ggplot(data, aes(x=x, y=y)) + 
  geom_point()
ggsave("filename.png", dpi=300, height=4, width=5, units="in")
```

# 2 Basics  
## 2.1 ggplot base figures    
You say you want to create a publication quality figure using `ggplot`? 
Well then, you are in the right place!  
First, let's look at what `ggplot` produces as a base graphic. 
We will use the R built in data set `mtcars`.

```{r base,eval=TRUE,fig.align='center',fig.cap="Figure 1. Base plot." }
library(ggplot2)
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point()
```

We get a gray background with small black dots, white grid lines, and a font that is not particularly legible when the figure is placed in this document. 
Most journals require a figure to be 300 dpi or greater not to mention the background should be white, without grid lines, and with a font that is consistent with the rest of the manuscript. 
Lets address these step by step.

## 2.2 White background  
Want a white background? We've got that. 
Simply add a `theme` to the figure. 
In this case we will add `theme_bw()`.  

```{r white, eval=TRUE,fig.align='center',fig.cap="Figure 1b. White background."}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point() + 
  theme_bw()
```

## 2.3 Remove grid lines  
To remove gridlines add another `theme`, and use element_blank() to clear the lines.  

```{r grids, eval=TRUE,fig.align='center',fig.cap="Figure 1c. White background, no gridlines."}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
   theme_bw() + 
   theme(panel.grid.major = element_blank(),
         panel.grid.minor = element_blank())
```

## 2.4 Fonts and resolution  
Fonts are both easy and not so easy to deal with depending on what your needs are.  
In this case we want a Times New Roman 12 pt font. 
The 12 pt is easy so let's start with that.
Simply add a base size option in `theme_bw()`.  

```{r font,eval=TRUE,fig.align='center',fig.cap="Fig 2. White background, no gridlines, plus a font size change."}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point() +
   theme_bw(base_size = 12) +
   theme(panel.grid.major = element_blank(),
         panel.grid.minor = element_blank())
```

Hey nothing happened, Figure 2 is the same as Figure 1c?  
The reason for this is the file size. 
These figures would be saved as .png and R has a base figure output of 72 ppi. 
This resolution is too low for journal submissions so let's up it. 
In this case I'm outputting a .png file with a set width and height and a resolution of 300. 

```{r save1, echo=F, dpi=300,fig.align='center',fig.cap="Fig 2b. Changed font size and resolution."}
ggplot(mtcars, aes(wt, mpg)) + geom_point()+
   theme_bw(base_size = 12) +
   theme(panel.grid.major = element_blank(),
         panel.grid.minor = element_blank())
```

In order to save the figure we will use `ggsave`. 
(Note the size increase of Figure 2b is due to presenting this on the web at 300 dpi - the ggsave function shown below will save a figure in a specified format at a chosen resolution and size).  

```{r save, eval=F}
ggsave("figure2b.png", dpi=300, dev='png', height=4.5, width=6.5, units="in")
```

Our figure is looking ok, but the font is not correct if you wanted *Time New Roman*. 
I'm on a Windows machine, so these procedures may be different for other operating systems. 
R is not terribly great at fonts so it is necessary to define the fonts clearly. 
This involves loading the `extrafont` package, then importing and unpacking the fonts. 
*Note: Only `font_import` one time, it takes a while and once done you are good to go, I have it commented out as I've already run it.*

```{r fontfake, message=FALSE, warning=FALSE,  eval=TRUE}
library(extrafont)
#font_import()
loadfonts(device = "win")
``` 

Load the fonts for a windows device and define the font for windows.

```{r loadwin, message=FALSE, warning=FALSE,  eval=TRUE}
windowsFonts(Times = windowsFont("TT Times New Roman"))
```

Now the font can be incorporated into the figure via the `theme_bw()` component (Figure 3).

```{r fig3, eval=T, dpi=300,fig.align='center',fig.cap="Fig 3. 12 pt Time New Roman font figure."}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point() +
  theme_bw(base_size=12, base_family='Times New Roman') + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

## 2.5 A better way: theme_set()

While all these adjustments to ggplot are great, there is a much better way for defining all of your
ggfigures. 
It goes by the name `theme_set`. 
Here is a working example.

First load the fonts (you've already run the `font_import()` command).
```{r, eval=F}
library(extrafont)
library(ggplot2)
loadfonts(device = "win")
windowsFonts(Times = windowsFont("TT Times New Roman"))
```

then use `theme_set` to determine the desired font and figure parameters.

```{r}
theme_set(theme_bw(base_size=12, base_family='Times New Roman')+
             theme(panel.grid.major = element_blank(), 
                   panel.grid.minor = element_blank()))
```

With the basic figure options set, less time is spent coding figures and more time can be spent
exploring data. The code used to produce Figure 3 is now reduced to:

```{r, eval=F}
ggplot(mtcars, aes(wt, mpg)) + geom_point()
```

# 3 Exploratory data analysis (very brief)  
## 3.1 Smoothers  

Lets explore these data, starting with a smooth. 
The base smooth for ggplot is *loess* (Figure 4) for data groups <1000 and a *gam* for groups >1000.

```{r, message=FALSE, eval=T, fig.align='center', fig.cap="Fig 4. Loess smoother."}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point() + 
  geom_smooth()
```

That is great and all but most of us do not use *loess* on a regular basis. 
However we can also implement other model structures (Figure 5).  

In this case we added a linear regression (lm) and generalized linear regression with a polynomial (glm), a generalized additive model with knots (*k*) set at 5, and a non-linear least squares model. 
*Note that the polynomial in the glm smooth could also be done via the lm smooth (also with the poly function instead of the identity component)*, however I'm mostly showing that there is a glm function that can be assigned different distributions
(e.g., family="binomial"), same goes for the gam smoother. 

More info can be found here https://stats.idre.ucla.edu/r/faq/how-can-i-explore-different-smooths-in-ggplot2/. 
Other good things can be found on that site
as well!

```{r, message=FALSE, eval=T,fig.align='center', fig.cap="Fig 5. Multiple smoothers."}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point() +   
   geom_smooth(alpha=.1) +    # loess smoother
   geom_smooth(method='lm', alpha=.1, color=2, fill=2) +    # linear model
   geom_smooth(method='glm', alpha=.1, color=3, fill=3, formula=y~x+I(x^2)) +    # generalized linear model
   geom_smooth(method='gam', alpha=.1, color=4, fill=4, formula=y~s(x, k=5)) +   # generalized additive model
   geom_smooth(method = "nls", formula = y ~ a*x^2 + b * x + c,  # non-linear least squares
               method.args = list(start=list(a=.1,b=.5,c=.2)), 
               se = FALSE, linetype = 1,
               colour = 5, fill=5, alpha=.1)
```

# 4. Plot options  
## 4.1 Point size & color  
Want bigger points, no problem simply change the size in `geom_point()` (Figure). 
Jitter using `geom_jitter()`, etc.

```{r point, message=FALSE,warning=FALSE,error=FALSE,  fig.align='center', fig.cap='Figure 6. Point size.'}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(size=4) + 
  geom_smooth(alpha=.1)
```


However, if you want the points to be different colors this can be added to the aesthetic component (in either the `ggplot()` or `geom_point()` component). 
For example lets change the size and color of the points by a car's horsepower (hp) a continuous variable in the dataframe. 

```{r color, message=FALSE,warning=FALSE,error=FALSE, fig.align='center', fig.cap='Figure 7. Continuous horsepower for color and size.'}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(size=hp, color=hp)) +
  geom_smooth(alpha=.1)
```

The colors and sizes could also be determined by changing hp to a factor.

```{r, message=FALSE,warning=FALSE,error=FALSE, fig.align='center', fig.cap='Figure 8. Factored horse power for point color and size.'}
library(tidyverse)
mtcars %>% 
  mutate(hp = factor(hp)) %>% 
  ggplot(aes(wt, mpg)) + 
  geom_point(aes(size = hp, color=hp)) +
  geom_smooth(alpha=.1)
```

However this will often create too many discrete groups to clearly understand what is being plotted, therefore a "better" method is to define the scales for the data points. This can be done with `scale_color_gradientn()`.

```{r, message=FALSE,warning=FALSE,error=FALSE, fig.align='center', fig.cap='Figure 9. Continuous horsepower color gradient.'}
mtcars %>% 
ggplot(aes(wt, mpg)) + 
  geom_point(aes(size=hp, color=hp))+
  geom_smooth(alpha=.1) +
  scale_colour_gradientn(colours=rainbow(5)) +
  scale_size(range=c(0,7))
```

Want to define where the legend breaks? no problem, just add a vector of values.

```{r, message=FALSE,warning=FALSE,error=FALSE, fig.align='center', fig.cap='Figure 10. Continuous horsepower color gradient with defined breaks.'}
bs <- seq(0,350,25) # define break locations

ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(size=hp, color=hp)) +
  geom_smooth(alpha=.1) +
  scale_colour_gradientn(colours=rainbow(5), breaks=as.vector(bs)) +
  scale_size(range=c(0,7), breaks=as.vector(bs))
```

Note that the legend may fall off the plotted area of the figure. 
One way to fix this is to remove one of the legends, in this case the legend that references the size of the points.

```{r, message=FALSE,warning=FALSE,error=FALSE, fig.align='center', fig.cap='Figure 11. Continuous horsepower color gradient with defined breaks and one legend removed.'}

ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(size=hp, color=hp)) +
  geom_smooth(alpha=.1) +
  scale_colour_gradientn(colours=rainbow(5), breaks=as.vector(bs)) +
  scale_size(range=c(0,7), breaks=as.vector(bs), guide = FALSE)
```

The legend now breaks in increments of 25 using a 5 color scale from the rainbow color palette.

## 4.2 Axis tickmarks  

It is quite easy to change the axis tick marks. Change the y-axis to every 5th value.

```{r tick,message=FALSE,warning=FALSE,error=FALSE, fig.align='center', fig.cap='Figure 12. Change y-axis tick marks.'}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(size=hp, color=hp)) +
  geom_smooth(alpha=.1) +
  scale_colour_gradientn(colours=rainbow(5), breaks=as.vector(bs)) +
  scale_size(range=c(0,7), breaks=as.vector(bs), guide = FALSE) +
  scale_y_continuous(breaks= seq(0, 40, 5))
```


Also change the x-axis, this time to unequal spacings.
```{r tick2,message=FALSE,warning=FALSE,error=FALSE, fig.align='center', fig.cap='Figure 13.Change y-axis and x-axis tick marks.'}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(size=hp, color=hp)) +
  geom_smooth(alpha=.1) +
  scale_colour_gradientn(colours=rainbow(5), breaks=as.vector(bs)) +
  scale_size(range=c(0,7), breaks=as.vector(bs), guide = FALSE) +
  scale_y_continuous(breaks= seq(0, 40, 5)) +
  scale_x_continuous(breaks= c(0, 1, 1.5, 2, 3, 5))
```

Your collaborator really prefers to have all of the tick marks present so... 
Change the labels while keeping the tickmarks.
```{r tick3,message=FALSE,warning=FALSE,error=FALSE, fig.align='center', fig.cap='Figure 14.Change y-axis and x-axis tick marks.'}
ggplot(mtcars, aes(wt, mpg)) + geom_point(aes(size=hp, color=hp)) +
  geom_smooth(alpha = 0.1) +
  scale_colour_gradientn(colours=rainbow(5), breaks=as.vector(bs)) +
  scale_size(range=c(0, 7), breaks=as.vector(bs),guide = FALSE) +
  scale_y_continuous(breaks= seq(0,40,5), labels=c(0,"",10,"",20,"",30,"",40)) +
  scale_x_continuous(breaks= seq(0,5,.5), labels=c(seq(0,2,.5),"","",3.5, "",4.5,""))
```

However, the scale presented doesn't include zero, something that is often nice to have. 
Most people approach this by using xlim() and/or ylim().

```{r zero,echo=TRUE, fig.align='center', fig.cap='Figure 15. Change y-axis and x-axis tick marks.',warning=FALSE}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(size=hp, color=hp)) +
  geom_smooth(alpha=.1) +
  scale_colour_gradientn(colours=rainbow(5), breaks=as.vector(bs)) +
  scale_size(range=c(0,7), breaks=as.vector(bs), guide = FALSE) +
  xlim(0,5) + ylim(0,40)
```

A number of things have now changed 

   - (1) the scales have reverted back to their original designations, 
   - (2) zero is now included on both axes, and 
   - (3) we have cut off data by constraining the axis at 5. 

One function of ggplot is that it only evaluates the data that is in the figure pane. 
So any smooths, etc., are informed only by the data that is seen. 
This can be changed to include the data that inform a smooth, but allow the figure to be "zoomed in" on an area of interest. 
To do this we use`coord_cartesian()`.

```{r cartesian,echo=TRUE, fig.align='center', fig.cap='Figure 16. Using coord_cartesian to "zoom in".'}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(size=hp, color=hp)) +
  geom_smooth(alpha=.1) +
  scale_colour_gradientn(colours=rainbow(5), breaks=as.vector(bs)) +
  scale_size(range=c(0,7), breaks=as.vector(bs), guide = FALSE) +
  scale_y_continuous(breaks= seq(0,40,5), labels=c(0,"",10,"",20,"",30,"",40)) +
  scale_x_continuous(breaks= seq(0,5,.5), labels=c(seq(0,2,.5),"","",3.5, "",4.5,"")) +
  coord_cartesian(xlim=c(0,5), ylim=c(0,40))
```

Or if you simply want to have the axes include zero ou could use `expand_limits`.

```{r cartesian2,echo=TRUE, fig.align='center', fig.cap= 'Figure 16a. Include zero using "expand_limits".'}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(size=hp, color=hp)) +
  geom_smooth(alpha=.1) +
  scale_colour_gradientn(colours=rainbow(5), breaks=as.vector(bs)) +
  scale_size(range=c(0,7), breaks=as.vector(bs), guide = FALSE) +
  expand_limits(x = 0, y = 0)
```

The smooth now extends to the end of the pane as it indeed includes the values that are off the page. 
The points can be brought back into the figure by changing the x-axis limits.

```{r xlimchange,echo=TRUE, fig.align='center', fig.cap='Figure 17. Include zero and show all data.'}

ggplot(mtcars, aes(wt, mpg)) +
  geom_point(aes(size=hp, color=hp)) +
  geom_smooth(alpha=.1) +
  scale_colour_gradientn(colours=rainbow(5), breaks=as.vector(bs)) +
  scale_size(range=c(0,7), breaks=as.vector(bs),guide = FALSE) +
  scale_y_continuous(breaks= seq(0,40,5), labels=c(0,"",10,"",20,"",30,"",40)) +
  scale_x_continuous(breaks= seq(0,5,.5), labels=c(seq(0,2,.5),"","",3.5, "",4.5,"")) +
  coord_cartesian(xlim=c(0,5.5), ylim=c(0,40))
```

# 5 Maps
## 5.1 Basic maps  

There are a number of ways in R to plot maps and deal with raster images etc. I have
no idea how to operate most of them. 
I do know a couple of easy ways to generate quick images though, they are presented below. 

**Going deeper into the mapping realm  
**(or the rabbit hole...)

Mapping can get very confusing very quickly depending upon what exactly you want to do,
how you want to present information, etc. 
I am focusing on a limited subset of mapping, mainly generating maps for quick visualization of data for evaluating spatial components. 
There is a great deal of information on mapping be sure to have a look around the interwebs. 
Let us generate a map of the western Gulf of Alaska.

```{r map1, fig.height=4, fig.cap="Figure 18. A map of Alaska, based on data from PBSmapping package."}
library(ggplot2)
library(mapdata)
ak <- map_data('worldHires','USA:Alaska')
ggplot() + 
  geom_polygon(data=ak, aes(long, lat, group=group), fill=8, color="black")
```


Notice that this map in Figure 18 doesn't seem quite right, there are a few ways to deal
with this. 
If you don't need the far western Aleutian Islands you can simply constrain
the plot (that is the method I will present); two, if you do need the Aleutian Islands then
use `world2Hires` as opposed to 'worldHires' for the data source, this centers the map on
the Pacific Ocean, but you will then have to deal with latitude and longitude adjustments
(longitude is no longer negative $\leftarrow$ small rabbit hole).

Lets return to the map we do want. 
There are two ways to constrain the data for the
Aleutians (remove it, or simply constrain the limits of the plot).
The first method here removes any points that are further west (east?) than 180$^o$. 
Then you can create a map with a light grey fill for the land and a black outline for the coast (Figure 19).

```{r map2, message=FALSE, fig.height=5, fig.cap='Figure 19. ggplot2 map of Alaska based upon worldHires data.', eval = T}
ak <- map_data('worldHires','USA:Alaska')
ak <- subset(ak, long<0) #drop the end of the Aleutian Islands, or use world2Hires 
ggplot() + 
  geom_polygon(data=ak, aes(long, lat, group=group), fill=8, color="black")
```


Or we can simply constrain the data for the map if we are only interested in a particular
area of Alaska. We can also clean the map up while zooming into a particular areas (Figure 24),
in this case the western Gulf of Alaska. Note that the code to zoom in is not simply done by
changing the x and y limits. When you constrain the x and y limits in ggplot it throws out
any data outside of the image, therefore the polygon shapes ("groups") will not be complete
and the image will not be pretty. In this plot I've incorporated the limits in
a projection wrapper (coord\_map), there are a number of different projections
that you can use though this will lead to a BIG rabbit hole!
I've added a background color (aliceblue) as well as incorporated degree symbols into the
x- and y-axis labels. Data points and polygons can now be added as additional layers as
shown earlier.
```{r map3, message=FALSE, fig.height=4, fig.cap='Figure 20. ggplot2 map of Alaska all dressed up with color.', fig.pos='H',dpi=300,dev='png', eval = T}

ggplot() + 
  geom_polygon(data = ak, aes(long, lat, group=group), fill = 8, color="black") +
  theme(panel.background = element_rect(fill = 'aliceblue')) +
  xlab(expression(paste(Longitude^o,~'W'))) +
  ylab(expression(paste(Latitude^o,~'N'))) +
  coord_map(xlim = c(-165, -145), ylim = c(54, 61))
```

 Figure 21 is the same map without color (better for reports, etc.).
```{r map4,message=FALSE, fig.height=4, fig.cap='Figure 21. Zoomed in ggplot map of the western Gulf of Alaska based upon PBSmapping data.',dpi=300,dev='png', fig.pos='H'}
ggplot() + 
  geom_polygon(data = ak, aes(long, lat, group = group), fill=8, color='black') +
   theme(panel.background = element_rect(fill = 'white')) + 
   xlab(expression(paste(Longitude^o,~'W'))) +
   ylab(expression(paste(Latitude^o,~'W'))) +
    coord_map(xlim = c(-165, -145), ylim = c(54, 61)) 
```

 Figure 22 is the same map without color but based upon a different base map layer. `PBSmapping` has a better defined coastline for Alaska than the `worldHires` database.

*Note:* `rnaturalearth` *is another good base map data set to consider using*. 

```{r mappbs,message=FALSE, fig.height=4, fig.cap='Figure 22. Zoomed in ggplot map of the western Gulf of Alaska based upon PBSmapping data.',dpi=300,dev='png', fig.pos='H'}
library(PBSmapping)
library(tidyverse)
data('nepacLLhigh') 
nepacLLhigh %>% 
  dplyr::select(group=PID, POS=POS,long=X, lat=Y) -> ak 

ggplot() + 
  geom_polygon(data = ak, aes(long, lat, group = group), fill=8, color='black') +
   theme(panel.background = element_rect(fill = 'white')) + 
   xlab(expression(paste(Longitude^o,~'W'))) +
   ylab(expression(paste(Latitude^o,~'W')))+
    coord_map(xlim = c(-165, -145), ylim = c(54, 61)) 
```

So you don't like the projection do you? 
Well change it using `+ coord_map()`. 
There are loads of different projections that can be used, here I show "gilbert", but there are plenty more. 
Look at page 4 here for various types https://cran.r-project.org/web/packages/mapproj/mapproj.pdf. 

```{r map5, message=FALSE, fig.height=5, fig.cap='Figure 23. ggplot2 map of Alaska based upon PBSmapping data with a "Gilbert" projection', eval = T}
library(mapproj)

ggplot() + 
  geom_polygon(data = ak, aes(long, lat, group = group), fill = 8, color = "black") +
  coord_map("gilbert")
```

## 5.3 Have shapefile - will map

The federal EEZ shapefile is available here https://github.com/ben-williams/excellent_figures/tree/master/shpfiles

This will require using `rgdal` and `rgeos` to manipulate the shapefile into a dataframe available for using in `ggplot2`. 
I've also used `ggrepel` as a great way to adjust labels in figures. 

`funcr` is my own package of miscellaneous functions. 
It is available here 
`install.packages("devtools")
devtools::install_github("ben-williams/funcr")`

Within it has `theme_report()` and `tickr()` functions to facilitate figure making.


```{r, fig.height=5, fig.cap='Figure 24. ggplot2 shapefile map of Alaska with federal management areas and a "Gilbert" projection', eval = T}
library(PBSmapping)
library(tidyverse)
library(ggrepel)
library(rgeos)
library(rgdal)
library(maptools)
library(funcr)
theme_set(theme_report())

# load the shapefile
areas <- readOGR("shpfiles", layer = "NMFS_GOA_WGS1984")
areas@data$id = rownames(areas@data)
areas.points = broom::tidy(areas, region="id")
areas.df = left_join(areas.points, areas@data, by="id")

#load PBSmapping data set for N Pacific - much better resolution than world highres...
data('nepacLLhigh') 
nepacLLhigh %>% 
  dplyr::select(group=PID, POS=POS,long=X,lat=Y) -> ak 

ports <- data.frame(long = c(-152.433333, -160.493333, -162.318056, -165.775), 
                    lat = c(57.787, 55.336667, 55.072222, 54.1325), 
                    port = c("Kodiak", "Sand Point", "King Cove", "Akutan"))



ggplot() +
  geom_polygon(data=ak, aes(long, lat, group=group), fill="lightgray", color="lightgray") + 
  coord_map("gilbert", xlim = c(-170, -145), ylim = c(49, 62)) +
  xlab(expression(paste(Longitude^o, ~'W'))) +
  ylab(expression(paste(Latitude^o, ~'N'))) +
  geom_path(data=areas.df, aes(long, lat, group = group)) +
  geom_text_repel(data = ports, aes(long, lat, label = port, group = port), point.padding = 1.27, direction = "y") +
  annotate("text", x = -165, y = 51.5, label = "Area \n 610") +
  annotate("text", x = -157, y = 53, label = "Area \n 620") +
  annotate("text", x = -151, y = 55, label = "Area \n 630") +
  annotate("text", x = -146, y = 57.2, label = "Area \n 640") +
  annotate("text", x = -167, y = 57.5, label = "Bering Sea") +
  annotate("text", x = -159, y = 60.5, label = "Alaska") +
  annotate("text", x = -150, y = 51, label = "Gulf of Alaska") 
```


## 5.4 Bathymetry maps

You need bathymetry data in order to make these. 
Can be provided as shapefiles or pulled from websources. 
Here are a few poorly explained examples.

```{r, fig.height=5, fig.cap='Figure 25. ggplot2 map of Alaska based upon PBSmapping and NOAA bathymetry data with a "Gilbert" projection', eval = T}
library(marmap)
library(inlmisc)


#  Download data from the server
getNOAA.bathy(lon1 = -175, lon2 = -150, lat1 = 53, lat2 = 66, resolution=1) -> AKBath
AKBath_df <- Grid2Polygons(as.SpatialGridDataFrame(AKBath), level=TRUE, pretty=TRUE)
AKBath_map <- broom::tidy(AKBath_df) %>% 
  mutate(id = factor(id, levels = paste0("X", 1:26)))

# function for a color gradient - note the white is under the map
colfunc <- colorRampPalette(c("royalblue4", "gray88", "white"))

ggplot() + 
  coord_map("gilbert", xlim = c(-165, -151), ylim = c(54, 61)) + 
  geom_map(data=AKBath_map, map=AKBath_map,
                    aes(map_id=id, fill = id), color="black") +
  geom_polygon(data = ak, aes(long, lat, group=group), fill=8, color="black") + 
  scale_fill_manual(values = colfunc(26)) + 
  theme(legend.position = "none") +
  scale_x_continuous(name = expression(paste(Longitude^o, ~'W')),
                     breaks = funcr::tickr(AKBath_map, long, to=1, start = -165, end = -150)$breaks,
                     labels = funcr::tickr(AKBath_map, long, to=1, start = -165, end = -150)$labels) +
  scale_y_continuous(name = expression(paste(Latitude^o, ~'N')),
                     breaks = funcr::tickr(AKBath_map, lat, to=1, start = 54)$breaks,
                     labels = funcr::tickr(AKBath_map, lat, to=1, start = 54)$labels)
```

Additional plotting options
```{r, cache=TRUE}
#  Plot bathy data
plot(AKBath, image=TRUE, deep=-6000, shallow=0, step=1000)
scaleBathy(AKBath, deg=2, x="topleft", inset=5)
```


```{r, cache=TRUE}
autoplot.bathy(AKBath, geom=c("r", "c")) + 
  scale_fill_etopo()
```
